### 2014-10-08 Pero_Transcriptome.tex
accepted!

The raw sequence reads corresponding to the four tissue types were error corrected using the software bless version 0.17 \citep{Heo:2014cb} using kmer=25, based on the developer's default recommendations (\url{https://github.com/macmanes/pero_transcriptome/blob/master/analyses.md#error-correction}). The error-corrected sequence reads were adapter and quality trimmed following recommendations from MacManes \citep{MacManes:2014io} and Mbandi \citep{Christoffels:2014bg}. Specifically, adapter sequence contamination and low quality nucleotides (defined as \textsc{Phred} \textless 2) were removed using the program Trimmomatic version 0.32  \citep{Bolger:2014ek}. Reads from each tissue were assembled using the Trinity version released 17 July 2014 \citep{Haas:2013jq}. We used flags to indicate the stranded nature of sequencing reads and set the maximum allowable physical distance between read pairs to 999nt (\url{https://github.com/macmanes/pero_transcriptome/blob/master/analyses.md#trinity-assemblies}). We elected to assembly reads derived from a single deeply sequenced individual (Peer360, a male) to reduce polymorphism and thus the complexity of the de Bruijn graph, which has important implications for runtime, hardware requirements \citep{Lowe:2014iv,Pop:2009dp}, and assembly contiguity \citep{Vijay:2012gy}. Individual tissues were assembled independently, as we hypothesize that tissue specific isoforms would be reconstructed with higher fidelity that if all tissues were assembled together.

The assembly was conducted on a linux workstation with 64 cores and 512Gb RAM. To filter the raw sequence assembly, we downloaded \textit{Mus musculus} cDNA and ncRNA datasets from Ensembl (\url{ftp://ftp.ensembl.org/pub/release-75/fasta/mus_musculus/}) and the \textit{Peromyscus maniculatus} reference transcriptome from NCBI (\url{ftp://ftp.ncbi.nlm.nih.gov/genomes/Peromyscus_maniculatus_bairdii/RNA/}). We used a blastN (version 2.2.29+) procedure (default settings, evalue set to 10$^{-10}$) to identify contigs in the \textit{P. eremicus} dataset likely to be biological in origin (\url{https://github.com/macmanes/pero_transcriptome/blob/master/analyses.md#blasting}). This procedure, when a reference dataset is available, retains more putative transcripts that a strategy employing expression-based filtering (remove if transcripts per million (TPM) \textless 1  \citep{MacManes:2012bu}) of the raw assembly. We then concatenated the filtered assemblies from each tissue into a single file and reduced redundancy using the software cd-hit-est version 4.6 \citep{Li:2006hr} using default settings, except that sequences were clustered based on 95\% sequence similarity (\url{https://github.com/macmanes/pero_transcriptome/blob/master/analyses.md#cd-hit-est}). This multi-fasta file was used for all subsequent analyses, including annotation and mapping. \\

@@

RNA was extracted from the hypothalamus, renal medulla, testes, and liver from each individual using sterile technique. TRIzol extraction resulted in a large amount of high quality (RIN $\geq$ 8) total RNA, which was then used as input. Libraries were constructed as per the standard Illumina protocol and sequenced as described above. The number of reads per library varied from 56 million strand-specific paired-end reads in Peer360 kidney, to 9 million single-end reads in Peer321 (\hypertarget{Table 1}{Table 1}, available as part BioProject PRJNA242486). Adapter sequence contamination and low-quality nucleotides were eliminated, which resulted in a loss of \textless 2\% of the total number of reads. These trimmed reads served as input for all downstream analyses. \\
\textbf{\hypertarget{Table 1}{Table 1}} \\
\begin{center}
\begin{tabular}{l|c | r}
\textsc{Dataset}& \textsc{Num. Raw Reads} & \textsc{SRA Accession} \\
\hline 
\textsc{Peer360 Testes} & 32M PE/SS  & SRR1575398 \\
\textsc{Peer360 Liver} & 53M PE/SS & SRR1575397 \\
\textsc{Peer360 Kidney} & 56M PE/SS  & SRR1575396 \\
\textsc{Peer360 Brain}  & 23M PE/SS & SRR1575395\\
\textsc{Peer305}  & 19M PE  & SRR1575434  \\
\textsc{Peer308}  & 15M PE  & SRR1575437 \\
\textsc{Peer319}  & 14M PE  & SRR1575439\\
\textsc{Peer321}  & 9M SE  & SRR1575441 \\
\textsc{Peer340}  & 16M PE  & SRR1575443 \\
\textsc{Peer352}  & 14M PE  & SRR1575464 \\
\textsc{Peer354}  & 9M SE  & SRR1575466 \\
\textsc{Peer359}  & 14M PE  & SRR1575492 \\
\textsc{Peer365}  & 16M PE & SRR1575493 \\
\textsc{Peer366}  & 16M PE & SRR1575494 \\
\textsc{Peer368}  & 14M PE  & SRR1575624\\
\textsc{Peer369}  & 14M PE  & SRR1575625\\
\textsc{Peer372}  & 17M SE  & SRR1576070\\
\textsc{Peer373}  & 23M SE  & SRR1576071\\
\textsc{Peer380}  & 16M SE  & SRR1576072 \\
\textsc{Peer382}  & 14M SE  & SRR1576073 \\
\end{tabular}
\begin{quote}
\small{Table 1. The number of sequencing reads per sample, whose identity is indicated by Peer[number]. PE=paired end, SS=strand specific, SE=single end sequencing. }
\end{quote}
\end{center}

Transcriptome assemblies for each tissue type was accomplished using the program Trinity \citep{Haas:2013jq}. The raw assemblies for brain, liver, testes, and kidney contained 185425, 222096, 180233, and 514091 assembled sequences respectively. This assembly was filtered using a blastN procedure against the \textit{Mus} cDNA and ncRNA and \textit{P. maniculatus} cDNAs, which resulted in a final dataset containing 68331 brain-derived transcripts, 71041 liver-derived transcripts, 67340 testes-derived transcripts, and 113050 kidney-derived transcripts. Mapping the error-corrected adapter/quality trimmed reads to these datasets resulted in mapping 94.98\% (87.01\% properly paired) of the brain-derived reads to the brain transcriptome, 96.07\% (88.13\% properly paired) of the liver-derived reads to the liver transcriptome, 96.81\% (85.10\% properly paired) of the testes-derived reads to the testes transcriptome, and 91.87\% (83.77\% properly paired) of the kidney-derived reads to the kidney transcriptome. Together, these statistics suggest that the tissue-specific transcriptomes are of extremely high quality. All tissue-specific assemblies are to be made available on Dryad, and until then are stored on Dropbox (\url{https://www.dropbox.com/sh/2jwzd8p6n6eluco/AAAO3nSdXb_u4wtQZRTwqW9ia?dl=0}). \\

@@

We then estimated gene expression on each of these tissue-specific datasets, which allowed us to understand expression patterns in the multiple tissues (Pero.tissue.xprs, will be made available on Dryad, until then on Dropbox (\url{https://www.dropbox.com/sh/2jwzd8p6n6eluco/AAAO3nSdXb_u4wtQZRTwqW9ia?dl=0})). Specifically, we constructed a Venn diagram ({\hypertarget{Figure 1}{Figure 1}}) that allowed us to visualize the proportion of genes whose expression was limited to a single tissue and those whose expression was ubiquitous. 66324 transcripts are expressed on all tissue types, while 13086 are uniquely expressed in the kidney, 2222 in the testes, 3580 in the brain, and 2798 in the liver. The kidney appears to an outlier in the number of unique sequences, though this could be the result of the recovery of more lowly expressed transcripts or isoforms.  \\

@@

To begin to test the hypothesis that selection on transcripts related to osmoregulation is enhanced in the desert adapted \peer, we calculated Tajima's D as described above, and implemented the branch-site test using alignments produced in MACSE. These alignments were manually inspected, and were relatively free from indels and internal stop codons. We set the sequence corresponding to \peer\:for Slc2a9, Vdr, and several of the Aquaporin genes (Aqp1,2,4,9) as the foreground lineages in six distinct program executions. These transcripts Slc2a9 and Vdr were chosen specifically because they - the former significantly - were recently linked to osmoregulation in a desert rodent \citep{Marra:2014de}. The test for Slc2a9 was highly significant (2$\Delta$Lnl=51.4, df=1, p=0, Table 5), indicating enhanced selection in \peer\: relative to the other lineages. The branch site test for positive selection conducted on the Vdr and Aquaporin genes were non-significant. While the branch site test of positive selection is largely non-significant, estimating Tajima's D for these few candidate loci demonstrates that either a selective or demographic process may be influencing the genome at these functionally relevant sites.  \\

>>>>

The raw sequence reads corresponding to the four tissue types were error corrected using the software bless version 0.17 \citep{Heo:2014cb} using kmer=25, based on the developer's default recommendations (\url{https://github.com/macmanes/pero_transcriptome/blob/master/analyses.md#error-correction}). The error-corrected sequence reads were adapter and quality trimmed following recommendations from MacManes \citep{MacManes:2014io} and Mbandi \citep{Christoffels:2014bg}. Specifically, adapter sequence contamination and low quality nucleotides (defined as \textsc{Phred} \textless 2) were removed using the program Trimmomatic version 0.32  \citep{Bolger:2014ek}. Reads from each tissue were assembled using the Trinity version released 17 July 2014 \citep{Haas:2013jq}. We used flags to indicate the stranded nature of sequencing reads and set the maximum allowable physical distance between read pairs to 999nt (\url{https://github.com/macmanes/pero_transcriptome/blob/master/analyses.md#trinity-assemblies}). We elected to assemble reads derived from a single deeply sequenced individual (Peer360, a male) to reduce polymorphism and thus the complexity of the de Bruijn graph, which has important implications for runtime, hardware requirements \citep{Lowe:2014iv,Pop:2009dp}, and assembly contiguity \citep{Vijay:2012gy}. Individual tissues were assembled independently, as we hypothesize that tissue specific isoforms would be reconstructed with higher fidelity than if all tissues were assembled together.

The assembly was conducted on a linux workstation with 64 cores and 512Gb RAM. To filter the raw sequence assembly, we downloaded \textit{Mus musculus} cDNA and ncRNA datasets from Ensembl (\url{ftp://ftp.ensembl.org/pub/release-75/fasta/mus_musculus/}) and the \textit{Peromyscus maniculatus} reference transcriptome from NCBI (\url{ftp://ftp.ncbi.nlm.nih.gov/genomes/Peromyscus_maniculatus_bairdii/RNA/}). We used a blastN (version 2.2.29+) procedure (default settings, evalue set to 10$^{-10}$) to identify contigs in the \textit{P. eremicus} dataset likely to be biological in origin (\url{https://github.com/macmanes/pero_transcriptome/blob/master/analyses.md#blasting}). This procedure, when a reference dataset is available, retains more putative transcripts than a strategy employing expression-based filtering (remove if transcripts per million (TPM) \textless 1  \citep{MacManes:2012bu}) of the raw assembly. We then concatenated the filtered assemblies from each tissue into a single file and reduced redundancy using the software cd-hit-est version 4.6 \citep{Li:2006hr} using default settings, except that sequences were clustered based on 95\% sequence similarity (\url{https://github.com/macmanes/pero_transcriptome/blob/master/analyses.md#cd-hit-est}). This multi-fasta file was used for all subsequent analyses, including annotation and mapping. \\

@@

RNA was extracted from the hypothalamus, renal medulla, testes, and liver from each individual using sterile technique. TRIzol extraction resulted in a large amount of high quality (RIN $\geq$ 8) total RNA, which was then used as input. Libraries were constructed as per the standard Illumina protocol and sequenced as described above. The number of reads per library varied from 56 million strand-specific paired-end reads in Peer360 kidney, to 9 million single-end reads in Peer321 (\hypertarget{Table 1}{Table 1}, available as part of BioProject PRJNA242486). Adapter sequence contamination and low-quality nucleotides were eliminated, which resulted in a loss of \textless 2\% of the total number of reads. These trimmed reads served as input for all downstream analyses. \\
\textbf{\hypertarget{Table 1}{Table 1}} \\
\begin{center}
\begin{tabular}{l|c | r}
\textsc{Dataset}& \textsc{Num. Raw Reads} & \textsc{SRA Accession} \\
\hline 
\textsc{Peer360 Testes} & 32M PE/SS  & SRR1575398 \\
\textsc{Peer360 Liver} & 53M PE/SS & SRR1575397 \\
\textsc{Peer360 Kidney} & 56M PE/SS  & SRR1575396 \\
\textsc{Peer360 Brain}  & 23M PE/SS & SRR1575395\\
\textsc{Peer305}  & 19M PE  & SRR1575434  \\
\textsc{Peer308}  & 15M PE  & SRR1575437 \\
\textsc{Peer319}  & 14M PE  & SRR1575439\\
\textsc{Peer321}  & 9M SE  & SRR1575441 \\
\textsc{Peer340}  & 16M PE  & SRR1575443 \\
\textsc{Peer352}  & 14M PE  & SRR1575464 \\
\textsc{Peer354}  & 9M SE  & SRR1575466 \\
\textsc{Peer359}  & 14M PE  & SRR1575492 \\
\textsc{Peer365}  & 16M PE & SRR1575493 \\
\textsc{Peer366}  & 16M PE & SRR1575494 \\
\textsc{Peer368}  & 14M PE  & SRR1575624\\
\textsc{Peer369}  & 14M PE  & SRR1575625\\
\textsc{Peer372}  & 17M SE  & SRR1576070\\
\textsc{Peer373}  & 23M SE  & SRR1576071\\
\textsc{Peer380}  & 16M SE  & SRR1576072 \\
\textsc{Peer382}  & 14M SE  & SRR1576073 \\
\end{tabular}
\begin{quote}
\small{Table 1. The number of sequencing reads per sample, whose identity is indicated by Peer[number]. PE=paired end, SS=strand specific, SE=single end sequencing. }
\end{quote}
\end{center}

Transcriptome assemblies for each tissue type were accomplished using the program Trinity \citep{Haas:2013jq}. The raw assemblies for brain, liver, testes, and kidney contained 185425, 222096, 180233, and 514091 assembled sequences respectively. This assembly was filtered using a blastN procedure against the \textit{Mus} cDNA and ncRNA and \textit{P. maniculatus} cDNAs, which resulted in a final dataset containing 68331 brain-derived transcripts, 71041 liver-derived transcripts, 67340 testes-derived transcripts, and 113050 kidney-derived transcripts. Mapping the error-corrected adapter/quality trimmed reads to these datasets resulted in mapping 94.98\% (87.01\% properly paired) of the brain-derived reads to the brain transcriptome, 96.07\% (88.13\% properly paired) of the liver-derived reads to the liver transcriptome, 96.81\% (85.10\% properly paired) of the testes-derived reads to the testes transcriptome, and 91.87\% (83.77\% properly paired) of the kidney-derived reads to the kidney transcriptome. Together, these statistics suggest that the tissue-specific transcriptomes are of extremely high quality. All tissue-specific assemblies are to be made available on Dryad, and until then are stored on Dropbox (\url{https://www.dropbox.com/sh/2jwzd8p6n6eluco/AAAO3nSdXb_u4wtQZRTwqW9ia?dl=0}). \\

@@

We then estimated gene expression on each of these tissue-specific datasets, which allowed us to understand expression patterns in the multiple tissues (Pero.tissue.xprs, will be made available on Dryad, until then on Dropbox (\url{https://www.dropbox.com/sh/2jwzd8p6n6eluco/AAAO3nSdXb_u4wtQZRTwqW9ia?dl=0})). Specifically, we constructed a Venn diagram ({\hypertarget{Figure 1}{Figure 1}}) that allowed us to visualize the proportion of genes whose expression was limited to a single tissue and those whose expression was ubiquitous. 66324 transcripts are expressed in all tissue types, while 13086 are uniquely expressed in the kidney, 2222 in the testes, 3580 in the brain, and 2798 in the liver. The kidney appears to an outlier in the number of unique sequences, though this could be the result of the recovery of more lowly expressed transcripts or isoforms.  \\

@@

To begin to test the hypothesis that selection on transcripts related to osmoregulation is enhanced in the desert adapted \peer, we calculated Tajima's D as described above, and implemented the branch-site test using alignments produced in MACSE. These alignments were manually inspected, and were relatively free from indels and internal stop codons. We set the sequence corresponding to \peer\:for Slc2a9, Vdr, and several of the Aquaporin genes (Aqp1,2,4,9) as the foreground lineages in six distinct program executions. The transcripts Slc2a9 and Vdr were chosen specifically because they - the former significantly - were recently linked to osmoregulation in a desert rodent \citep{Marra:2014de}. The test for Slc2a9 was highly significant (2$\Delta$Lnl=51.4, df=1, p=0, Table 5), indicating enhanced selection in \peer\: relative to the other lineages. The branch site tests for positive selection conducted on the Vdr and Aquaporin genes were non-significant. While the branch site test of positive selection is largely non-significant, estimating Tajima's D for these few candidate loci demonstrates that either a selective or demographic process may be influencing the genome at these functionally relevant sites.  \\
